---
title: Data frame operations
author: Robert McDonald
bibliography: ../refs.bib
---


There are typically several different ways to accomplish a specific
task in R. The purpose of this document is to illustrate different
ways to undertake operations on data frames. Even if you prefer one
method and stick to it, the odds are that you will encounter examples
and co-workers who have different preferences. 


Here are the libraries we will use:

```{r, echo=TRUE, message=FALSE}
library(knitr)
library(tidyverse)
opts_chunk$set(collapse=TRUE,
               message=FALSE,
               comment=NA
               )
```

We begin by defining a data frame, named `df`:^[This examples was
inspired by @ruhlman:09. Ratios are by weight. Any errors are mine.]
```{r}
df <- data.frame(what=c('crust', 'cookie', 'cake', 'pasta'),
                 flour=c(3, 3, 1, 3),
                 butter=c(2, 2, 1, 0),
                 liquid=c(1, 0, 0, 0),
                 sugar=c(0, 1, 1, 0),
                 egg=c(0, 0, 1, 2),
                 stringsAsFactors=FALSE
                 )

dft <- tibble(what=c('crust', 'cookie', 'cake', 'pasta'),
              flour=c(3, 3, 1, 3),
              butter=c(2, 2, 1, 0),
              liquid=c(1, 0, 0, 0),
              sugar=c(0, 1, 1, 0),
              egg=c(0, 0, 1, 2)
              )
all.equal(df, dft, check.attributes=FALSE)
```
The `data.frame` function will automatically create strings as factors
unless we tell it not to. Many would prefer character vectors not
being treated as factors, so this can be an advantage of the `tibble`
function.

The `all.equal` function will complain that `df` and `dft` have
different classes unless we specify `check.attributes=FALSE`.

The data frame looks like this:
```{r}
kable(df)
```

Data frame operations include: 

* adding one or more new column
* deleting one or more columns
* adding one or more new rows
* deleting one or more rows
* renaming columns

We will illustrate ways to do these in base R and in `dplyr`.

## Extract a column

Suppose we want to extract the flour column and have it be a vector.

```{r}
c1 <- df[, 'flour']
c2 <- df$flour
c3 <- df[['flour']]
c4 <- df['flour'][[1]]
class(c1)  ## this is a vector
all.equal(c1, c2)
all.equal(c1, c3)
all.equal(c1, c4)
```


Suppose we want to extract the `flour` column and keep it as a
dataframe (a list).

```{r}
c5 <- df['flour']  
c6 <- df %>% select(flour)
class(c5)  ## this is a dataframe
all.equal(c5, c6)
all.equal(c1, c5)  ## a vector and dataframe are not the same!
```


## Add a column

We will add "nuts" and "yeast" columns:

```{r}
df1 <- df2 <- df3 <- df  ## create copies of the original data frame
df1$nuts <- c(0, 0.5, 0, 0)
df1$yeast <- 0  ## the recycling rule in action
df2[, "nuts"] <-  c(0, 0.5, 0, 0)
df2[, "yeast"] <- 0
df3 <- df %>% mutate(nuts=c(0, 0.5, 0, 0), yeast=0)
all.equal(df1, df2)
all.equal(df2, df3)
kable(df1)
```

Very important: Notice that when using `dplyr` to construct `df3`, no
quotes are needed. When using base R, quotes are needed for column
names. The absence of quotation within the `dplyr` universe is great,
but it can be very tricky (and frustrating) to mix `dplyr` and base
R. My advice is to stick with one or the other as much as possible.

## Delete columns

Now we decide we don't need the columns we just added:

```{r}
df4 <- df1;
df4[c('nuts', 'yeast')] <- NULL  ## assigning to `NULL` deletes an object
df5 <- df1[-c(7, 8)]
df6 <- df1[-which(names(df1) %in% c('nuts', 'yeast'))]
df7 <- df1[, -which(names(df1) %in% c('nuts', 'yeast'))]
df8 <- df1 %>% select(-nuts, -yeast)
all.equal(df4, df5)
all.equal(df4, df6)
all.equal(df4, df7)
all.equal(df4, df8)

```

## Add rows

We will use the dataframe including nut and yeast columns.

```{r}
newrow1 <- list('bread', 1, 0, .67, 0, 0, 0, .02)
newrow2 <- list(what='bread', flour=1, butter=0,
                liquid=.67, sugar=0, egg=0, nuts=0, yeast=.02)
df.ar1 <- rbind(df1, newrow1)
df.ar2 <- rbind(df1, newrow2)
df.ar3 <- bind_rows(df1, newrow2) ## dplyr
all.equal(df.ar1, df.ar2)
all.equal(df.ar1, df.ar3)
```
## Filter rows

We can choose to keep or delete rows that meet specific criteria. We
will use `df.ar1` as the base dataframe.

Suppose we want only items that use butter

```{r}
df.butter1 <- df.ar1[df.ar1$butter > 0, ]
df.butter2 <- df.ar1[df.ar1$"butter" > 0, ]
df.butter3 <- subset(df.ar1, df.ar1$butter > 0)
df.butter4 <- df.ar1 %>% filter(butter > 0)
all.equal(df.butter1, df.butter2)
all.equal(df.butter1, df.butter3)
all.equal(df.butter1, df.butter4)
```
Note that any expression evaluating to a logical will work. So
compound conditions (e.g. `df.ar1$butter > 0 & df.ar1$liquid > 0`)
will work fine.

# References
