---
title: Comparing Rnw to Rmd
subtitle: An example
author: 
- Robert McDonald\thanks{Kellogg School of Management, Northwestern University}
- Joe Smith\thanks{Someother University}
date: "`r format(Sys.Date(), '%B %e, %Y')`"
bibliography: ../refs.bib
latex-engine: xelatex
thanks: Kellogg School of Management, Northwestern University 
abstract: "It's possible to create a document using either 
\\LaTeX\ \ coupled with R code, or using Rmarkdown coupled with R code. 
How hard is it to do one vs the other? This is the Rmarkdown version."
output: 
  pdf_document:
    number_sections: true
geometry: margin=1.4in
fontsize: 12pt
toc: true
colorlinks: true
linkcolor: 'red'
toccolor: 'cyan'
linktocpages: true
number_sections: true
header-includes: 
   - \usepackage{titletoc}
   - \usepackage{appendix}
---

\thispagestyle{empty}

\newpage

```{r init, echo=FALSE, message=FALSE}
library(tidyverse)
library(babynames)
library(knitr)
library(kableExtra)
opts_chunk$set(message=FALSE,
               comment=NA)
names <- c('Mary', 'James')
print(getwd())
```

# Introduction 

This is a test document to illustrate the difference between Rmd and 
Rnw files. 

## Auto-references and equations

It is possible to include equations (such as \autoref{eq:test}  below)
and references (such as to @grolemund/wickham:17 or @schmuller:17)

\begin{equation}
\alpha + \beta/\Delta + \int_0^\infty f(x) dx
\label{eq:test}
\end{equation}


# Babynames 

## Top 10 names historically

\autoref{tbl:top10} shows the top names historically, from 1880 to 2015.

```{r toptable, echo=FALSE}
topnames <- babynames %>%
    group_by(name) %>%
    summarize(Total=sum(n)) %>%
    arrange(-Total) %>%
    filter(row_number() <= 10)

tbl <- left_join(topnames, babynames) %>% 
    group_by(name, Total) %>%
    summarize('Year most popular'=as.character(year[which.max(prop)]),
              'Proportion that year'=max(prop)) %>%
    arrange(-Total) %>%
    rename(Name=name) %>%
    kable(caption='Top 10 names, 1880-2015\\label{tbl:top10}',
          digits=4,
          format.args=list(big.mark=','),
          align='lccc'
          ) 
kable_styling(tbl, "latex")
          
```

## Popularity of `r paste(names, collapse=' and ')`

\autoref{fig:babyfig} shows the popularity of the names `r#
paste(names, collapse=' and ')` over `r#
length(unique(babynames$year))` years. The placement of the figure is
tricky. Latex permits the placement options "hptb". When placing the
figure, knitr has to decide whether to write a figure environment or a
markdown figure specification (`![]()`). The way to force a figure
environment (from [this post on
stackoverflow](https://stackoverflow.com/questions/42486617/knitr-ignoring-fig-pos))
is to use a figure caption and at least one of fig.align, out.width,
or out.extra. It works to use `out.extra=''`, which is what I do
in this document.


```{r babyfig, message=FALSE, echo=FALSE, fig.cap='Popularity of James and Mary from 1880 to 2015\\label{fig:babyfig}', fig.height=4, out.extra='', fig.pos='p'}
babynames %>%
    filter(name %in% c('James', 'Mary')) %>%
    ggplot(aes(x=year, y=prop, color=sex)) +
    geom_point() +
    facet_wrap( ~ name, scales='fixed') + theme_bw()
```

\newpage

\appendixpage
\appendix

# Code

## Table code

Here is the code used to produce \autoref{tbl:top10}:

```{r, eval=FALSE}
<<toptable>>
```

## Figure code

Here is the code used to produce \autoref{fig:babyfig}:

```{r, eval=FALSE}
<<init>>
<<babyfig>>
```


\newpage

# References
